Semantic Linearity in a Lazy Optimising Compiler

### 2024-01-29 icfp.tex
passed through seciton 5

\todo[inline]{Todo este paragrafo parece-me redundante}
Resources used in a scrutinee that is not in WHNF must
not be used in the case alternatives.
%
%
However, it is not sufficient to evaluate the scrutinee to weak head normal
form to \emph{fully} consume all resources used in the scrutinee, since
sub-expressions such as constructor arguments will be left unevaluated. To
\emph{fully} consume all resources occurring in the scrutinee, the scrutinee
must be evaluated either to \emph{normal form} or such that all linear components of
the scrutinee are fully evaluated, as witnessed by the $Alt0$ rule above. In short,
for a case expression whose scrutinee is not in WHNF, the scrutinee
resources must \emph{not} be used directly in the case alternatives,
but the result of evaluating the scrutinee to WHNF must still be
consumed, as all sub-expressions of the scrutinee remain unevaluated and must
be consumed. Since the scrutinee resources cannot be consumed directly, they must be
consumed indirectly through $\D$-variables, namely, either the case binder, or
the linear pattern-bound variables introduced in the alternative.
% We tackle this in due time, in the proof irrelevance section.
% \begin{itemize}

@@

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Mapsto
\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
% Essentially, if the scrutinee is in WHNF the matching alternative is easily
% determined, and must be treated with the specialized $\Mapsto$ judgement, which
% only applies to the matching constructor.
% %
Alternatives not matching the scrutinee, as mentioned in the discussion of the
$Alt0$ rule, could use resources arbitrarily as they will never be executed,
however, we uniformly treat non-matching alternatives as if the scrutinee were
not in WHNF. Having introduced proof irrelevant resources, we can now
present the updated and fully general $Case_\textrm{WHNF}$ rule:
\[
\TypeCaseWHNF
\]
We note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely
written by humans.
Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
Crucially, type preservation for the case-reduction substituting the
case binder and pattern variables in the alternative is not possible
to prove without branching on WHNF-ness, since otherwise the
$\D$-substitution is not well-defined.  \todo[inline]{What is the
  $\Delta$-substitution? Clarify}

@@

We have implemented the Linear Core type system as a plugin for the Glasgow
Haskell Compiler.
GHC plugins allow developers to inspect and modify programs being compiled by
GHC, at different stages of compilation~\cite{10.1145/3331545.3342599}.
%
In particular, for any given Haskell module, Core plugins run for (and receive
as input) every intermediate program produced in the compilation process, i.e.
from desugaring and after each optimising transformation.
%
This implementation further substantiates our claim that Linear Core is
suitable for the intermediate language of an optimising compiler.

The GHC Linear Core plugin~\cite{cite:linear-core-plugin} implements a
typechecker for Linear Core: given a Core program, our plugin typechecks the
linearity of all expressions bound in that program, failing if a linear
resource is not used \emph{exactly once} according to semantic linearity of the
$(\lambda^\pi_\Delta)$ system.
%
Most notably, the plugin successfully validates linearity throughout the
(optimising) compilation of linearity-heavy libraries, namely
\texttt{linear-base} and \texttt{linear-smc}, except in expressions whose
linearity depends on so-called \emph{multiplicity coercions}, which are an
avenue of future work ($\S$~\ref{sec:future-work}) exploring the intersection
of Linear Core with type equality coercions.
% Furthermore, the implementation accepts all example programs from Chapter~\ref{sec:linearity-semantically}
% deemed well-typed by Linear Core, which are marked with a
% \colorbox{notyet}{\notyetcolorname} background.

@@

The implementation of Linear Core as a typechecker does not follow directly
from the description of the type system because Linear Core is not
\emph{syntax-directed}. Specifically, the most challenging features are
splitting linear resources amongst sub-derivations and consuming fragments of
resources through pattern-bound variables.
%
We use the following techniques to tackle the \emph{non-syntax-directedness} of
Linear Core, thus making the system more suitable to implement:
%
\begin{itemize}

\item Instead of non-deterministically splitting linear resources amongst
sub-derivations, we thread input/output linear resources through each
derivation using the resource management for linear logic described
by~\cite{DBLP:journals/tcs/CervesatoHP00}.
% \todo{Escrever uma regra?}

% In a case expression whose scrutinee is not in WHNF,
\item Pattern variables bound in a case alternative, for a scrutinee not in WHNF,
are introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
variable. To use the resources through the pattern-bound $\D$-vars, they must
be first $Split$ into fragments.

We consume tagged fragments of a resource \emph{as needed}, i.e., when a resource,
whose usage environment has a fragmented resource with tag $K_j$, is used, we $Split$
the matching resource according to the constructor $K$ and consume the fragment
$K_j$, rather than eagerly determining which resources need to be fragmented to
do so.
%
We note that it is safe to destructively fragment the resource, i.e. removing
the whole resource and only leaving the split fragments, because resources are
only $Split$ when a fragment is needed and, consequently, if a fragment is
consumed, using the ``whole'' resource as well violates linearity.

\end{itemize}
%
Furthermore, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments, but discuss an avenue of further research regarding
this inference challenge in Section~\ref{sec:future-work}.

@@

The results of running the Linear Core GHC plugin on large established
libraries focused around linear types are given by
Figure~\ref{fig:core-plugin-res}.
%
We compiled the libraries \texttt{linear-smc}, a library presented in the
work~\cite{10.1145/3471874.3472980} (1500 lines), \texttt{linear-base} (4000
lines), the Haskell standard library for programming with linear types,
comprised of over 100 modules, and
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
session-types library, using our plugin.
%
We count the number of programs accepted by our implementation, where each
top-level binding in a module counts as a program, and every such binding is
typechecked once per optimisation pass. i.e., we typecheck all intermediate
programs produced by GHC. The total amount of programs rejected by the
typechecker are given in the ``Total Rejected'' column, but we distil these
rejections into ``Unique Rejections'' by removing duplicate rejections (those
of the same program and for the same linearity-violating reason, but occurring
at different stages). We further categorize the unique rejections into ``Linear
modulo Call-by-name'', ``Linear Rejected'', ``Not Linear Rejected'', and
``Unknown Rejected''.
%
Using the plugin, \texttt{linear-base} takes 35 seconds to compile, instead of
20 seconds. We note, however, that the implementation is not performance
conscious, and types every intermediate program from scratch, instead of
maintaining linearity information throughout the pipeline.

Programs ``linear modulo call-by-name'' are a class of programs
which scrutinize a variable, but then uses the variable in the case
alternatives of the case scrutinizing it. As will be made clear in
Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
understood as linear as long as applications of linear functions binding these
variables are \emph{not} reduced \emph{call-by-name}, because, in doing so,
linear resources are duplicated. In Core, these programs are not seen as
linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
% We have a flag in our implementation to accept some of these programs...
%
``Linear Rejected'' programs include different kinds of programs which are
rejected by the Linear Core system, but are still semantically linear. An
example is a program to which a rewrite rule was applied, resulting in an
application of unrestricted function to a linear resource.
% in the first argument
% to \incode{build}:
% \begin{lstlisting}
% take :: Int -> Replicator a *' $\lolli$ '* [a]
% take = \ (ds :: Int) (r :: Replicator b) ->
%      case ds of
%        1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
%        ...
% \end{lstlisting}
% We know the linear resource is still used linearly because \incode{build} will
% always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
% linear. Additionally, these include programs scrutinizing a value of a type of
% which all constructors have no linear components, but only matching on the
% default alternative, programs where common-sub-expression elimination
% substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
% scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
%
``Not Linear Rejected'' indicate programs that we do not understand as linear,
semantically, and are simultaneously rejected by Linear Core and its
implementation. These programs were not necessarily identified as bad
by the linter.\footnote{rever}
% An example, where the last component of \incode{HashMap}, \incode{wwz},
% is linear, but is not being consumed in the case alternative matching on it:
% \begin{lstlisting}
% jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
% jssvi (a :: Ur Bool) (b :: Set Int)
%   = case a of
%      Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
% \end{lstlisting}
%
Finally, ``Unknown Rejected'' programs are those whose validity we did not
check. These include both programs accepted by Linear Core, but not by its
implementation, and programs that are simply rejected by Linear Core, but which
were not categorized.

\input{prototype/core-plugin-results}

>>>>

\todo[inline]{Todo este paragrafo que esta commented out parece-me redundante}
% Resources used in a scrutinee that is not in WHNF must
% not be used in the case alternatives.
% %
% %
% However, it is not sufficient to evaluate the scrutinee to weak head normal
% form to \emph{fully} consume all resources used in the scrutinee, since
% sub-expressions such as constructor arguments will be left unevaluated. To
% \emph{fully} consume all resources occurring in the scrutinee, the scrutinee
% must be evaluated either to \emph{normal form} or such that all linear components of
% the scrutinee are fully evaluated, as witnessed by the $Alt0$ rule above. In short,
% for a case expression whose scrutinee is not in WHNF, the scrutinee
% resources must \emph{not} be used directly in the case alternatives,
% but the result of evaluating the scrutinee to WHNF must still be
% consumed, as all sub-expressions of the scrutinee remain unevaluated and must
% be consumed. Since the scrutinee resources cannot be consumed directly, they must be
% consumed indirectly through $\D$-variables, namely, either the case binder, or
% the linear pattern-bound variables introduced in the alternative.
%ATE AQUI
% We tackle this in due time, in the proof irrelevance section.
% \begin{itemize}

@@

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Mapsto
\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
% Essentially, if the scrutinee is in WHNF the matching alternative is easily
% determined, and must be treated with the specialized $\Mapsto$ judgement, which
% only applies to the matching constructor.
% %
Alternatives not matching the scrutinee, as mentioned in the discussion of the
$Alt0$ rule, could use resources arbitrarily as they will never be executed,
however, we uniformly treat non-matching alternatives as if the scrutinee were
not in WHNF. Having introduced proof irrelevant resources, we can now
present the correct $Case_\textrm{WHNF}$ rule:
\[
\TypeCaseWHNF
\]
We note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely
written by humans.
Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
Crucially, type preservation for the case-reduction substituting the
case binder and pattern variables in the alternative is not possible
to prove without branching on WHNF-ness, since otherwise the
$\D$-substitution is not well-defined.  \todo[inline]{What is the
  $\Delta$-substitution? Clarify}

@@

We have implemented the Linear Core type system as a plugin for the Glasgow
Haskell Compiler.
GHC plugins allow developers to inspect and modify programs being compiled by
GHC, at different stages of compilation~\cite{10.1145/3331545.3342599}.
%
In particular, for any given Haskell module, Core plugins run for (and receive
as input) every intermediate program produced in the compilation process, i.e.
from desugaring and after each optimising transformation.
%
This implementation further substantiates our claim that Linear Core is
suitable for the intermediate language of an optimising compiler.
\todo[inline]{Tem de ser anonimizado, provavelmente alterando a bib
  entry para omitir autores e o proprio link}
The GHC Linear Core plugin~\cite{cite:linear-core-plugin} implements a
typechecker for Linear Core: given a Core program, our plugin typechecks the
linearity of all expressions bound in that program, failing if a linear
resource is not used \emph{exactly once} according to semantic linearity of the
$(\lambda^\pi_\Delta)$ system.
%
The plugin successfully validates linearity throughout the
(optimising) compilation of linearity-heavy libraries, namely
\texttt{linear-base} and \texttt{linear-smc}, except in expressions whose
linearity depends on so-called \emph{multiplicity coercions}, which are an
avenue of future work (Section~\ref{sec:future-work}) exploring the intersection
of Linear Core with type equality coercions.
% Furthermore, the implementation accepts all example programs from Chapter~\ref{sec:linearity-semantically}
% deemed well-typed by Linear Core, which are marked with a
% \colorbox{notyet}{\notyetcolorname} background.

@@

The implementation of Linear Core as a typechecker does not follow
directly from the description of the type system because Linear Core
is not \emph{syntax-directed}. Specifically, the most challenging
features are splitting linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
Instead of non-deterministically splitting linear resources amongst
sub-derivations, we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.  Pattern variables
bound in a case alternative for a scrutinee not in WHNF, are
introduced as $\D$-variables with usage environment
$\lctag{\irr{\D}}{K_i}$, where $\D$ are the scrutinee resources and
$K_i$ the tag of that pattern variable. To use the resources through
the pattern-bound $\D$-vars, they must be first $Split$ into
fragments. We consume tagged fragments of a resource \emph{as needed},
i.e., when a resource, whose usage environment has a fragmented
resource with tag $K_j$, is used, we $Split$ the matching resource
according to the constructor $K$ and consume the fragment $K_j$,
rather than eagerly determining which resources need to be fragmented
to do so.
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. Further research regarding
this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

% % \item Instead of non-deterministically splitting linear resources amongst
% % sub-derivations, we thread input/output linear resources through each
% % derivation using the resource management for linear logic described
% % by~\cite{DBLP:journals/tcs/CervesatoHP00}.
% % \todo{Escrever uma regra?}

% % In a case expression whose scrutinee is not in WHNF,
% \item Pattern variables bound in a case alternative, for a scrutinee not in WHNF,
% are introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
% where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
% variable. To use the resources through the pattern-bound $\D$-vars, they must
% be first $Split$ into fragments.

% We consume tagged fragments of a resource \emph{as needed}, i.e., when a resource,
% whose usage environment has a fragmented resource with tag $K_j$, is used, we $Split$
% the matching resource according to the constructor $K$ and consume the fragment
% $K_j$, rather than eagerly determining which resources need to be fragmented to
% do so.
% %
% We note that it is safe to destructively fragment the resource, i.e. removing
% the whole resource and only leaving the split fragments, because resources are
% only $Split$ when a fragment is needed and, consequently, if a fragment is
% consumed, using the ``whole'' resource as well violates linearity.

% \end{itemize}
%

@@

The results of running the Linear Core GHC plugin on large 
libraries focused around linear types are given by
Figure~\ref{fig:core-plugin-res}.
%
We compiled the libraries \texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines), \texttt{linear-base} (4000
lines), the Haskell standard library for programming with linear types,
comprised of over 100 modules, and
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
session-types library, using our plugin.
%
We count the number of programs accepted by our implementation, where
each top-level binding in a module counts as a program, and every such
binding is typechecked once per optimisation pass. i.e., we typecheck
all intermediate programs produced by GHC. The total amount of
programs rejected by the typechecker are given in the ``Total
Rejected'' column, but we distil these rejections into ``Unique
Rejections'' by removing duplicate rejections (those of the same
program and for the same linearity-violating reason, but occurring at
different stages). We further categorize the unique rejections into
``Linear modulo Call-by-name'', ``Linear Rejected'', ``Not Linear
Rejected'', and ``Unknown Rejected''.  We note that our implementation
is not performance conscious. For instance, \texttt{linear-base} takes
35 seconds to compile with our plugin, instead of 20 seconds.
%
% Using the plugin, \texttt{linear-base} takes 35 seconds to compile, instead of
% 20 seconds. We note, however, that the implementation is not performance
% conscious, and types every intermediate program from scratch, instead of
% maintaining linearity information throughout the pipeline.

Programs ``linear modulo call-by-name'' are a class of programs
which scrutinize a variable, but then use the variable in the case
alternatives of the case scrutinizing it. As discussed in
Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
understood as linear as long as applications of linear functions binding these
variables are \emph{not} reduced \emph{call-by-name}.
In Core, these programs are not seen as
linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
% We have a flag in our implementation to accept some of these programs...
%
``Linear Rejected'' programs include different kinds of programs which are
rejected by the Linear Core system, but are still semantically linear. An
example is a program to which a rewrite rule was applied, resulting in an
application of unrestricted function to a linear resource.
% in the first argument
% to \incode{build}:
% \begin{lstlisting}
% take :: Int -> Replicator a *' $\lolli$ '* [a]
% take = \ (ds :: Int) (r :: Replicator b) ->
%      case ds of
%        1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
%        ...
% \end{lstlisting}
% We know the linear resource is still used linearly because \incode{build} will
% always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
% linear. Additionally, these include programs scrutinizing a value of a type of
% which all constructors have no linear components, but only matching on the
% default alternative, programs where common-sub-expression elimination
% substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
% scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
%
``Not Linear Rejected'' indicate programs that we do not understand as linear,
semantically, and are simultaneously rejected by Linear Core and its
implementation. % These programs were not necessarily identified as bad
% by the linter.\footnote{rever}
% An example, where the last component of \incode{HashMap}, \incode{wwz},
% is linear, but is not being consumed in the case alternative matching on it:
% \begin{lstlisting}
% jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
% jssvi (a :: Ur Bool) (b :: Set Int)
%   = case a of
%      Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
% \end{lstlisting}
%
Finally, ``Unknown Rejected'' programs are those whose validity we did not
check. These include both programs accepted by Linear Core, but not by its
implementation, and programs that are simply rejected by Linear Core, but which
were not categorized.

{\small
  \input{prototype/core-plugin-results}
  }
